Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-05 paper.tex
Added in Joergs edits

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\textit{Tautomerism} - Tautomers are isomers of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\textit{Protonation} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites, geochemicals, and
petrochemicals. Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries, test hypotheses, and provides guidance in the
design of optimal combinatorial experiments.

@@

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
valuable resources to search for new drugs. For instance the well known
ligand-based design approach is making use of these databases along with
structure similarity tools to discover new compounds binding to biological
targets (protein or nucleic acid). Compounds retrieved through ligand-based
design, however, must already be present in databases, which is a limitation of
the technique. To palliate that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching circular [5]
or path fingerprints [6], these fingerprints being popular tools to measure
similarity. Enumerating molecules from fingerprints, or more generally
enumerating molecules from structure-activity molecular descriptors, is still
very much under development. At stake is the design of new chemicals or drugs
that have not yet been synthesized.

Another area where enumeration plays a role is in the \emph{generation of chemical
reaction networks}. The problem here consists of enumerating all the possible
compounds that can be produced by applying reaction rules to a set of initial
molecules. By reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of E.J. Corey, who received the
Nobel Prize in Chemistry in 1990. Designing new drugs or chemicals,
understanding the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and synthetic
biology to produce heterologous compounds in microorganisms, all are
applications that require the enumeration of reaction networks. As reviewed in Chapter
11 in [1] several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of intermediate
compounds being produced. One way to limit the number of compounds generated is
to simulate the dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
compute on-the-fly species concentrations. Chemical reaction network enumeration
and sampling is an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic engineering
strategies to biosynthesize compounds of commercial interest. The difficulty
with metabolic network design is that in addition to network generation based on
reactions, one also needs to verify that there are enzymatic sequences enabling
reactions catalyses. That additional task requires encompassing both chemical
and sequence information and the development of tools that are at the interface
between chemoinformatics and bioinformatics.

@@

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an industrial and academic discipline with roots dating back to the 1960s, with the creation of DENDRAL, the first expert system of any kind.  Much of its development from the 1960s until the new millenium was performed in secret, because of competition in the pharmaceutical industry.  While chemists were among the first scientists to embrace web-based tools of knowledge management, research into fundamental algorithmic questions of structure enumeration and database management (and a host of other topics) have suffered, because of limited academic access to either tools or data.  As we enter the second decade of the 21st century, however, this landscape is rapidly shifting.  There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens of millions of compounds.  We believe the prospects for academic and industrial collaboration between chemists and computer scientists are quite bright, and we would like to encourage more computer scientists to participate in cheminformatics research.  There is an admittedly steep learning curve, as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework.  Nevertheless, because of the changing landscape, there are many research questions we are only now in position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; for a database designer, more effective search algorithms and ontologies; for a practical programmer, expansion of the many open-source cheminformatics projects.  If more chemists think algorithmically, and more computer scientists think ``chemically,'' we may discover ``miracle drugs'' more consistently than ever before.

>>>>

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion. The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, bi-protonated, or any variations of those states).
\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites, for naming just one example. 
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries, test hypotheses, and provides guidance in the
design of optimal combinatorial experiments.

@@

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
valuable resources to search for new drugs. For instance the well known
ligand-based design approach is making use of these databases along with
structure similarity tools to discover new compounds binding to biological
targets (protein or nucleic acid). Compounds retrieved through ligand-based
design, however, must already be present in databases, which is a limitation of
the technique. To palliate that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching circular [5]
or path fingerprints [6], these fingerprints being popular tools to measure
similarity. Enumerating molecules from fingerprints, or more generally
enumerating molecules from molecular descriptors, is still
very much under development. At stake is the design of new chemicals or drugs
that have not yet been synthesized.

Another area where enumeration plays a role is in the \emph{generation of chemical
reaction networks}. The problem here consists of enumerating all the possible
compounds that can be produced by applying reaction rules to a set of initial
molecules. By reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of E.J. Corey, who received the
Nobel Prize in Chemistry in 1990. Designing new drugs or chemicals,
understanding the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and synthetic
biology to produce heterologous compounds in microorganisms, all are
applications that require the enumeration of reaction networks. As reviewed in Chapter
11 in [1] several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of intermediate
compounds being produced. One way to limit the number of compounds generated is
to simulate the dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
compute on-the-fly species concentrations. Chemical reaction network enumeration
and sampling is an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic engineering
strategies to biosynthesize compounds of commercial interest. The difficulty
with metabolic network design is that in addition to network generation based on
reactions, one also needs to verify that there are enzymatic events possible for enabling
reactions catalyses. That additional task requires encompassing both chemical
and sequence information and the development of tools that are at the interface
between chemoinformatics and bioinformatics.

@@

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an industrial 
and academic discipline with roots dating back to the 1960s, with the creation of 
DENDRAL, the first expert system of any kind.  Much of its development from the 1960s 
until the new millenium was performed in secret, because of competition in the pharmaceutical industry.  
While chemists were among the first scientists to embrace web-based tools of knowledge management, 
research into fundamental algorithmic questions of structure enumeration and database management 
(and a host of other topics) have suffered, because of limited academic access to either tools or data.  
As we enter the second decade of the 21st century, however, this landscape is rapidly shifting.  
There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens 
of millions of compounds.  We believe the prospects for academic and industrial collaboration between 
chemists and computer scientists are quite bright, and we would like to encourage more computer 
scientists to participate in cheminformatics research.  There is an admittedly steep learning curve, 
as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as
 ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework. 
Nevertheless, because of the changing landscape, there are many research questions we are only now in 
position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; 
for a database designer, more effective search algorithms and ontologies; for a practical programmer, 
expansion of the many open-source cheminformatics projects.  
If more chemists think algorithmically, and more computer scientists think ``chemically,'' 
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals into account (only one identifier and isomer possible per molecule), 
but also their complex relationships, transformations, 
and combinatorial challenges into account. This will allow us to normalize, search, and use all chemical information, 
around us and within us.
